---
title: "Report on Climate Change"
author: "Christopher Carbone"
date: "2/3/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This is a report on global climate change. First, we will examine carbon emission rates since the beginning of the industrial revolution compared with global temperature anomalies. Next, we will look at the atmospheric concentrations of the 3 main greenhouse gases over the last millennium. Finally, we will explore the historic levels of atmospheric CO~2~ over the last 800,000 years.

We will be using data sets loaded from the [dslabs package](https://cran.r-project.org/web/packages/dslabs/dslabs.pdf).

We start by loading the necessary libraries:

```{r loading-libs}
library(tidyverse)
library(dslabs)
```

then loading the first data set:

```{r loading-data-1}
data("temp_carbon")
```

## Carbon emissions and global temperature anomalies

Notice that after carbon emissions begin to rise at the end of the 19th century, global temperature anomalies soon follow suit with an upward trend:

```{r carbon-emissions-and-temp-anomalies, echo=FALSE}
temp_carbon %>% ggplot(aes(year)) +
  geom_line(aes(y=carbon_emissions/1000)) +
  ylab("carbon emissions in billions of metric tons") +
  ggtitle("Annual carbon emissions by year, 1751-2014")

temp_carbon %>% ggplot(aes(year)) +
  geom_abline(slope=0, col="gray") +
  geom_line(aes(y=temp_anomaly, col="combined anomaly"), alpha=.5, lwd=.5) +
  geom_line(aes(y=land_anomaly, col="land anomaly"), alpha=.5, lwd=.5) +
  geom_line(aes(y=ocean_anomaly, col="ocean anomaly"), alpha=.5, lwd=.5) +
  geom_text(x=1810, y=0.06, label="20th Century Mean", col="gray65") +
  ylab(expression("temperature anomaly ("*~degree*"C)")) +
  ggtitle("Temperature anomaly relative to 20th century mean, 1880-2018") +
  theme(legend.position = c(.2,.7))
```

***
We load the next data set:

```{r loading-data-2}
data("greenhouse_gases")
```

## Greenhouse gas concentrations

We see a relatively stable plateau spanning centuries leading up to an exponential increase in atmospheric concentrations of CH~4~, CO~2~, and N~2~O beginning in 1850, marking the start of widespread fossil fuel consumption in the industrial revolution:

```{r greenhouse-gas-concentrations, echo=FALSE}
greenhouse_gases %>% ggplot(aes(year, concentration)) +
  geom_line() +
  facet_grid(gas~., scales="free") +
  geom_vline(xintercept = 1850, col="gray65") +
  scale_x_continuous(breaks = c(0,500,1000,1500,1850,2000)) +
  ylab(expression(paste("Concentration (",CH[4],"/",N[2],"O ppb, ",CO[2]," ppm)"))) +
  ggtitle("Atmospheric greenhouse gas concentration by year, 0-2000")
```

***
We load the final data set:

```{r loading-data-3}
data("historic_co2")
```

## Historic CO~2~ atmospheric concentration

We see natural cycles of CO~2~ increase and decrease lasting 50,000 to 100,000 years per cycle, with the decrease happening more slowly than the increase, and a sharp spike upward in recent history:

```{r historic-co2, echo=FALSE}
co2_time <- historic_co2 %>% ggplot(aes(year/1000, co2, col=source)) +
  geom_line() +
  ylab(expression(paste(CO[2]," concentration in ppm by volume"))) +
  xlab("millenium")

co2_time + 
  ggtitle(expression(paste("Atmospheric ",CO[2]," concentration over 800,000 years")))
```

Zooming in on the period around 790,000 years ago, we see that it took roughly 10,000 years for CO~2~ levels to rise from 200 ppmv to its peak around 265 ppmv:

```{r historic-co2-790000-years-ago, echo=FALSE}
co2_time + xlim(-800, -775) +
  ggtitle(expression(paste("Atmospheric ",CO[2]," 800,000 BCE - 775,000 BCE")))
```

Looking more closely at the period around 350,000 years ago, we see that it took over 27,000 years for CO~2~ levels to rise from 180 ppmv to 300 ppmv:

```{r historic-co2-350000-years-ago, echo=FALSE}
co2_time + xlim(-375, -330) +
  ggtitle(expression(paste("Atmospheric ",CO[2]," 375,000 BCE - 330,000 BCE")))
```

We see in the period around 130,000 years ago that it took over 11,000 years for CO~2~ levels to rise from 195 ppmv to 285 ppmv:

```{r historic-co2-130000-years-ago, echo=FALSE}
co2_time + xlim(-140, -120) +
  ggtitle(expression(paste("Atmospheric ",CO[2]," 140,000 BCE - 120,000 BCE")))
```

Lastly, we see that in recent history, CO~2~ levels held a steady plateau for over 4,500 years before spiking from 275 ppmv to almost 410 ppmv in the last 250 years or so, with a sharp upturn right around 1850, once again coinciding with the start of widespread fossil fuel consumption:

```{r historic-co2-last-5000-years, echo=FALSE}
historic_co2 %>% ggplot(aes(year, co2, col=source)) +
  geom_line() +
  geom_vline(xintercept = 1850, col="gray65") +
  ylab(expression(paste(CO[2]," concentration in ppm by volume"))) +
  xlim(-3000, 2018) +
  geom_text(x=1420, y=225, label="1850 CE", col="gray65") +
  ggtitle(expression(paste("Atmospheric ",CO[2]," 3000 BCE - 2018 CE")))
```

***
## Conclusion

The data depicts a strong correlation between rising carbon emissions, atmospheric greenhouse gas levels, and global warming trends that perfectly coincide with the rise of widespread fossil fuel consumption around 1850 CE.

***
